---
title: "Interactively explore and visualize Single Cell RNA seq data"
author: "Jayaram Kancherla, Kazi Tasnim Zinat, Héctor Corrada Bravo"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Explore Data using TreeViz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

[`TreeViz`](https://github.com/HCBravoLab/TreeViz) is a package for interactive visualization and exploration of Single Cell RNA sequencing data. `TreeViz` provides methods for exploring hierarchical features (eg. clusters in single cell at different resolutions or taxonomic hierarchy in single cell datasets), while supporting other useful data visualization charts like heatmaps for expression and scatter plots for TSNE.

## Loading required packages

```{r load-packages, message=FALSE, warning=FALSE}
library(TreeViz)
library(dplyr)
library(Seurat)
library(SC3)
library(scran)
library(scater)
library(clustree)
library(igraph)
library(scRNAseq)
```

# Preparing Datasets

The first step in using the `Treeviz` package is to wrap datasets into `TreeViz` objects. The `TreeViz` class extends `SummarizedExperiment` and provides various methods to interactively perform various operations on the underlying hierarchy and count or expression matrices. In this section, we show various ways to generate a `TreeViz` object either from existing Single Cell packages (SingleCellExperiment or Seurat) or from a raw count matrix and cluster hierarchy.

## Create TreeViz from count matrix and Cluster hierarchy

```{r, results='hide', warning=FALSE, error=FALSE, message=FALSE}
n=64
# create a hierarchy
df<- data.frame(cluster0=rep(1,n))
for(i in seq(1,5)){
  df[[paste0("cluster",i)]]<- rep(seq(1:(2**i)),each=ceiling(n/(2**i)),len=n)
}

# generate a count matrix
counts <- matrix(rpois(6400, lambda = 10), ncol=n, nrow=100)

treeViz <- createTreeViz(df, counts)
plot(treeViz)
```

## From `Seurat`

We use the dataset `pbmc_small` available through Seurat to create a `TreeViz` object.  

```{r, eval=FALSE, echo=TRUE, results='hide', warning=FALSE, error=FALSE, message=FALSE}
data(pbmc_small)
pbmc <- pbmc_small
```

The measurements for dimensionality reduction methods we want to visualize are also added to the object via native functions in `Seurat`. Since `PCA` is already available, we calculate `TSNE` and `UMAP`

```{r, eval=FALSE, echo=TRUE, results='hide', warning=FALSE, error=FALSE, message=FALSE}
pbmc<- RunTSNE(pbmc)
pbmc<- RunUMAP(pbmc, dims=1:3)
Reductions(pbmc)
```

We provide the `createFromSeurat` function to create a `TreeViz` object from `Seurat` object. In addition, we can also provide the name of dimensionality reductions present in the object as a parameter to indicate these dimensions also be wrapped in the `treeviz` object for visualization. 

Note: If the reduced dimension is not present it would be ignored.

```{r, eval=FALSE, echo=TRUE,  warning=FALSE, error=FALSE, message=FALSE}
treeViz<- createFromSeurat(pbmc, reduced_dim = c("umap","pca","tsne"))
plot(treeViz)
```

## From `SingleCellExperiment`

A number of Single cell datasets are available as `SingleCellExperiment` objects through the `scRNAseq` package, for this usecase, we use `LunSpikeInData` dataset. In addition, we calculate the dimensionality reductions; UMAP, TSNE and PCA from the functions provided in `scater` package. 

```{r, results='hide', warning=FALSE, error=FALSE, message=FALSE}

Similar to the `Seurat` methods, we provide `createFromSCE` function to create a `TreeViz` object from `SingleCellExperiment` object. Here, the workflow works in two ways:

1. If no cluster information is available in the `colData` of the `SingleCellExperiment` object, we create clusters at different resolutions using the `WalkTrap` algorithm by calling an internal function `generate_walktrap_hierarchy` and use this cluster information for visualization.

2. If cluster information is provided in the `colData` of the object, then the user should set the flag parameter `check_colData` to `TRUE` and provide prefix for the columns where cluster information is stored.

Note: In both cases the user needs to provide the name of dimensionality reductions present in the object as a parameter.

```{r, eval=FALSE, warning=FALSE, error=FALSE, message=FALSE}
treeViz <- createFromSCE(sce, reduced_dim = c("UMAP","PCA","TSNE"))
plot(treeViz)
```

We create clusters at different resolutions using the `WalkTrap` algorithm

```{r, eval=TRUE, echo=TRUE, results='hide', warning=FALSE, error=FALSE, message=FALSE}
data_matrix <- as.matrix(logcounts(sce))

SNN_igraph = scran::buildKNNGraph(
  data_matrix, 
  transposed = FALSE)

snn.matrix = similarity(
      SNN_igraph, 
      method = "jaccard")
    
snn.graph <- graph_from_adjacency_matrix(snn.matrix, weighted = TRUE, mode = "undirected")

walktrapClusters = igraph::cluster_walktrap(snn.graph)
for (i in 1:10){
  sce[[paste0("cluster",i)]] <-  factor(igraph::cut_at(walktrapClusters, n = i))
}

treeviz<- createFromSCE(sce)

We provide `startTreeViz` method to initialize interactive exploration of data from the `treeViz` object we created before. 

```{r, eval=FALSE, echo=TRUE}
app <- startTreeviz(treeViz, top_genes = 500)
```

This process

```{r, warning=FALSE, error=FALSE, message=FALSE}
treeViz <- createFromSCE(sce)
plot(treeViz)
```

## Visualize gene expression across clusters

## The `TreeViz` session manager

Start the App from the `treeViz` object we created. This adds a `facetZoom` to navigate the cluster hierarchy, a heatmap of the top 100 most variable genes from the dataset and a scatter plot with TSNE dimensions. 
 
```{r, eval=FALSE, echo=TRUE}
app <- startTreeviz(treeViz)
```

Users can also use the interface to explore the same dataset using different visualizations available through Epiviz.

## Stop App

After exploring the dataset, this command the websocket connection. 

```{r, eval=FALSE, echo=TRUE}
app$stop_app()
```