---
title: "Interactively explore and visualize Single Cell RNA seq data"
author: "Kazi Tasnim Zinat, Jayaram Kancherla, Héctor Corrada Bravo"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Explore Data using TreeViz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

[`TreeViz`](https://github.com/HCBravoLab/TreeViz) is a package for interactive visualization and exploration of Single Cell RNA sequencing data. `TreeViz` provides methods for exploring hierarchical features (eg. clusters in single cell at different resolutions or taxonomic hierarchy in single cell datasets), while supporting other useful data visualization charts like heatmaps for expression and scatter plots for TSNE.

## loading required packages

```{r load-packages, message=FALSE, warning=FALSE}
library(TreeViz)
library(dplyr)
library(Seurat)
library(SC3)
library(scran)
library(scater)
library(clustree)
library(igraph)
```

# Preparing Datasets

In this section, we show ways to generate a `TreeViz` object from existing Single Cell packages (SingleCellExperiment or Seurat) or from a raw count matrix and cluster hierarchy

## Create TreeViz from count matrix and Cluster hierarchy

```{r, results='hide', warning=FALSE, error=FALSE, message=FALSE}
n=64
df<- data.frame(cluster0=rep(1,n))
for(i in seq(1,5)){
  df[[paste0("cluster",i)]]<- rep(seq(1:(2**i)),each=ceiling(n/(2**i)),len=n)
}
counts <- matrix(rpois(6400, lambda = 10), ncol=n, nrow=100)

treeViz <- createTreeViz(df, counts)
plot(treeViz)
```


## From `Seurat`

We use the dataset `pbmc_small` available through Seurat to create a `TreeViz` object.  

```{r, eval=TRUE, echo=TRUE, results='hide', warning=FALSE, error=FALSE, message=FALSE}
pbmc <- pbmc_small
```

We then preprocess the data and find clusters at different resolutions.

```{r, eval=FALSE, echo=TRUE, results='hide', warning=FALSE, error=FALSE, message=FALSE}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
pbmc <- NormalizeData(pbmc)
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, vars.to.regress = "percent.mt")
pbmc <- FindVariableFeatures(object = pbmc)
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0), print.output = 0, save.SNN = TRUE)
pbmc
```

We use the helper function to create a TreeViz object from Seurat

```{r, eval=TRUE, echo=TRUE,  warning=FALSE, error=FALSE, message=FALSE}
treeViz <- createFromSeurat(pbmc)
plot(treeViz)
```

## From `SingleCellExperiment`

We create a `mock SingleCellExperiment` object

```{r, results='hide', warning=FALSE, error=FALSE, message=FALSE}

sce<- mockSCE()

# Normalization.
sce <- logNormCounts(sce)

# define feature names in feature_symbol column
rowData(sce)$feature_symbol <- rownames(sce)
# remove features with duplicated names
sce <- sce[!duplicated(rowData(sce)$feature_symbol), ]

# Dimensionality reduction.
sce <- runPCA(sce)
```

We create clusters at different resolutions using the `WalkTrap` algorithm

```{r, eval=TRUE, echo=TRUE, results='hide', warning=FALSE, error=FALSE, message=FALSE}
data_matrix <- as.matrix(logcounts(sce))

SNN_igraph = scran::buildKNNGraph(
  data_matrix, 
  transposed = FALSE)

snn.matrix = similarity(
      SNN_igraph, 
      method = "jaccard")
    
snn.graph <- graph_from_adjacency_matrix(snn.matrix, weighted = TRUE, mode = "undirected")

walktrapClusters = igraph::cluster_walktrap(snn.graph)
for (i in 1:10){
  sce[[paste0("cluster",i)]] <-  factor(igraph::cut_at(walktrapClusters, n = i))
}

treeviz<- createFromSCE(sce)

sce
```

We use the helper function to create a `TreeViz` object from SCE

```{r, warning=FALSE, error=FALSE, message=FALSE}
treeViz <- createFromSCE(sce)
plot(treeViz)
```

# Interactively explore `TreeViz` object

## The `TreeViz` session manager

Start the App from the `treeViz` object we created. This adds a `facetZoom` to navigate the cluster hierarchy, a heatmap of the top 100 most variable genes from the dataset and a scatter plot with TSNE dimensions. 
 
```{r, eval=FALSE, echo=TRUE}
app <- startTreeviz(treeViz)
```

Users can also use the interface to explore the same dataset using different visualizations available through Epiviz.

## Stop App

After exploring the dataset, this command the websocket connection. 

```{r, eval=FALSE, echo=TRUE}
app$stop_app()
```
