---
title: "Interactively explore and visualize Single Cell RNA seq data"
author: "Jayaram Kancherla, Kazi Tasnim Zinat, Héctor Corrada Bravo"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Explore Data using TreeViz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

[`TreeViz`](https://github.com/HCBravoLab/TreeViz) is a package for interactive visualization and exploration of Single Cell RNA sequencing data. `TreeViz` provides methods for exploring hierarchical features (eg. clusters in single cell at different resolutions or taxonomic hierarchy in single cell datasets), while supporting other useful data visualization charts like heatmaps for expression and scatter plots for TSNE.

## Loading required packages

```{r load-packages, message=FALSE, warning=FALSE}
library(TreeViz)
library(dplyr)
library(Seurat)
library(SC3)
library(scran)
library(scater)
library(clustree)
library(igraph)
library(scRNAseq)
```

# Preparing Datasets

The first step in using the `Treeviz` package is to wrap datasets into `TreeViz` objects. The `TreeViz` class extends `SummarizedExperiment` and provides various methods to interactively perform various operations on the underlying hierarchy and count or expression matrices. In this section, we show various ways to generate a `TreeViz` object either from existing Single Cell packages (SingleCellExperiment or Seurat) or from a raw count matrix and cluster hierarchy.

## Create TreeViz from count matrix and Cluster hierarchy

```{r, results='hide', warning=FALSE, error=FALSE, message=FALSE}
n=64
# create a hierarchy
df<- data.frame(cluster0=rep(1,n))
for(i in seq(1,5)){
  df[[paste0("cluster",i)]]<- rep(seq(1:(2**i)),each=ceiling(n/(2**i)),len=n)
}

# generate a count matrix
counts <- matrix(rpois(6400, lambda = 10), ncol=n, nrow=100)

treeViz <- createTreeViz(df, counts)
plot(treeViz)
```

## From `Seurat`

We use the dataset `pbmc_small` available through Seurat to create a `TreeViz` object.  

```{r, eval=FALSE, echo=TRUE, results='hide', warning=FALSE, error=FALSE, message=FALSE}
data(pbmc_small)
pbmc <- pbmc_small
```

We then preprocess the data and find clusters at different resolutions.

```{r, eval=FALSE, echo=TRUE, results='hide', warning=FALSE, error=FALSE, message=FALSE}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
pbmc <- NormalizeData(pbmc)
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, vars.to.regress = "percent.mt")
pbmc <- FindVariableFeatures(object = pbmc)
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0), print.output = 0, save.SNN = TRUE)
pbmc
```

We use the helper function to create a TreeViz object from Seurat

Note: If the reduced dimension is not present it would be ignored.

```{r, eval=FALSE, echo=TRUE,  warning=FALSE, error=FALSE, message=FALSE}
treeViz<- createFromSeurat(pbmc, reduced_dim = c("umap","pca","tsne"))
plot(treeViz)
```

## From `SingleCellExperiment`

A number of Single cell datasets are available as `SingleCellExperiment` objects through the `scRNAseq` package, for this usecase, we use `LunSpikeInData` dataset. In addition, we calculate the dimensionality reductions; UMAP, TSNE and PCA from the functions provided in `scater` package. 

```{r, results='hide', warning=FALSE, error=FALSE, message=FALSE}

Similar to the `Seurat` methods, we provide `createFromSCE` function to create a `TreeViz` object from `SingleCellExperiment` object. Here, the workflow works in two ways:

1. If no cluster information is available in the `colData` of the `SingleCellExperiment` object, we create clusters at different resolutions using the `WalkTrap` algorithm by calling an internal function `generate_walktrap_hierarchy` and use this cluster information for visualization.

# define feature names in feature_symbol column
#rowData(sce)$feature_symbol <- rownames(sce)
# remove features with duplicated names
#sce <- sce[!duplicated(rowData(sce)$feature_symbol), ]

# Dimensionality reduction.
#sce <- runPCA(sce)
sce<- runTSNE(sce)
```

We create clusters at different resolutions using the `WalkTrap` algorithm

```{r, eval=FALSE, echo=TRUE, results='hide', warning=FALSE, error=FALSE, message=FALSE}
data_matrix <- as.matrix(logcounts(sce))

SNN_igraph = scran::buildKNNGraph(
  data_matrix, 
  transposed = FALSE)

snn.matrix = similarity(
      SNN_igraph, 
      method = "jaccard")
    
snn.graph <- graph_from_adjacency_matrix(snn.matrix, weighted = TRUE, mode = "undirected")

walktrapClusters = igraph::cluster_walktrap(snn.graph)
for (i in 1:10){
  sce[[paste0("cluster",i)]] <-  factor(igraph::cut_at(walktrapClusters, n = i))
}

treeviz<- createFromSCE(sce)

sce
sce<- generate_walktrap_hierarchy(sce)
```

This process

```{r, warning=FALSE, error=FALSE, message=FALSE}
treeViz <- createFromSCE(sce)
plot(treeViz)
```

## Visualize gene expression across clusters

## Start the TreeViz App (using hosted app)

Start the App from the `treeViz` object we created. This adds a `facetZoom` to navigate the cluster hierarchy, a heatmap of the top 100 most variable genes from the dataset and a scatter plot with TSNE dimensions. 
 
```{r, eval=FALSE, echo=TRUE}
app <- startTreeviz(treeViz)
```

Users can also use the interface to explore the same dataset using different visualizations available through Epiviz.

## Stop App

After exploring the dataset, this command the websocket connection. 

```{r, eval=FALSE, echo=TRUE}
app$stop_app()
```


## Start TreeViz with app running locally

```{r, eval=FALSE, echo=TRUE}
setTreevizStandalone()
app <- startTreevizStandalone(treeViz)

```


## Stop App

```{r, eval=FALSE, echo=TRUE}
app$stop_app()
```




