---
title: "Basic create and access TreeIndex objects"
author: "Jayaram Kancherla, Hector Corrada Bravo"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to metavizr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## loading required packages

```{r load-packages, message=FALSE, warning=FALSE}
library(TreeSummarizedExperiment)
library(metagenomeSeq)
library(msd16s)
```


## Example Data

```{r}
data(mouseData)
counts <- MRcounts(mouseData)
hierarchy <- fData(mouseData)
```


## Create `TreeIndex` object from feature hierarchy
```{r}
tree <- TreeIndex(hierarchy)
tree
```

## Operations on `TreeIndex` object

Similar to subset operations in data frame, `TreeIndex` objects can subset leaf nodes in the tree. The result is a new `TreeIndex` object with the filtered rows.

```{r}
subset <- tree[1:1000,]
subset
```

We can also prune the tree to a given level as shown below. The result is a list of nodes and their indices. The indices can then be used to aggregate counts in `SummarizedExperiment` objects

```{r}
tree_level <- splitAt(tree, selectedLevel = 2)
tree_level
```

We can also format the result of a prune operation as a `TreeIndex` object. however, the mapping of indices to the count matrix are lost in translation.

```{r}
tree_level <- splitAt(tree, selectedLevel = 2, format="TreeIndex")
tree_level
```
